Estimating the socio-environmental impacts of car substitution by bicycle and public transit using open tools

### 2023-09-26 paper/PaperTRA/PaperTRA.tex
paper TRA - translt
stopped at  section 2.4 (will resume soon)

@Robinlovelace can you take a look at the paragraph in the introduction about PCT?

\usepackage[hidelinks]{hyperref}

@@

\maketitle              % typeset the header of the contribution
%
\begin{abstract}
In metropolitan areas, car trips can be replaced by a combination of
public transit and cycling for the first-and-last mile. This paper
focuses on estimating the potential for cycling + PT as a substitute for
car trips in the Lisbon metropolitan area and assessing its
socio-environmental impacts using open data and open source tools. A
decision support tool that facilitates the design and development of a
metropolitan cycling network was developed (\emph{biclaR}). A scenario
of intermodality introduced, and its socio-environmental impacts were
assessed using the \emph{HEAT for Cycling} and the \emph{HEAT as a
Service} tools. Additionally, the impacts of shifting car trips to PT
were estimated and monetized. The results indicate that 20\% of the
current trips can be made with the bicycle + PT combination. Shifting to
cycling for the first-and-last mile can reduce annual
CO\textsubscript{2}eq emissions from 6,000 tons/day, and the 10-year
socio-environmental benefits account from €230 million. For the PT leg,
the transfer from car results in the avoidance of at lest 8,500 tons of
CO\textsubscript{2}eq emissions per year. The information on
socio-economic benefits can support policymakers in prioritizing
interventions to reduce the reliance on individual motorized
transportation and effectively communicate their decisions.

@@

\textbf{full paper}: 4-6 pages in length (typically up to 3,000 words).

In metropolitan areas, car trips can be replaced by a combination of
public transit (PT) and cycling for the first-and-last mile. This
approach requires interventions and programs to make bicycling more
appealing, and the resulting public investments can have significant
social and environmental benefits. This paper focuses on estimating the
potential for cycling + PT as a substitute for car trips in the Lisbon
metropolitan area (LMA) and assessing its socio-environmental impacts
using open data and open source tools.

According to the latest mobility survey conducted in 2018, the LMA
registered a total of 5.3 million daily trips, with only 0.5\% by
bicycle. Car modal share is 58.4\%, while PT accounts for 15.5\%. To
achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively), the Department
of Transport introduced biclaR, a decision support tool that facilitates
the design and development of a metropolitan cycling network.

This research aims to present and discuss the methods used to
estimate\ldots{}

Propensity to Cycle Tool

adding up an intermodality scenario to estimate cycling potential to
public transit interfaces, and thus to support planning and prioritize
investments in the cycling network.

@@

\hypertarget{case-study}{%
\subsection{Case Study}\label{case-study}}

As características das viagens do IMob, constituem o cenário base deste
projeto. Este inquérito à mobilidade foi realizado em 2017. Apesar de
ter sido realizado em período pré-pandemia, este conjunto de dados é a
melhor e mais recente informação que temos em termos de mobilidade
urbana nas áreas metropolitanas.

Segundo o IMob (2017), das cerca de 5.3 milhões de viagens diárias na
área metropolitana de Lisboa, apenas 25 479 das viagens são realizadas
em bicicleta (0,5\%), enquanto 3.1 milhões são feitas em automóvel
(58.4\%), 1.3 milhões a pé (23.9\%), 825 mil em transportes públicos
(15.5\%) e 96 mil em outros modos (1.8\%).

O número de viagens intramunicipais - ou que são realizadas com origem e
destino no mesmo município (3.5 milhões viagens) - é superior ao número
de viagens intermunicipais - com origem e destino em municípios
diferentes (1.8 milhões viagens). O automóvel e os transportes públicos
são os modos maioritariamente utilizados em viagens intermunicipais. O
automóvel é o modo mais utilizado em qualquer tipologia de viagem.

Este é o cenário base, ou de referência, utilizado apenas para
comparação com os cenários seguintes - de onde advém o potencial
ciclável.

@@

imob data

Aplicou-se um método de desagregação das origens e destinos das viagens
entre o centróide de uma freguesia para o centróide de outra freguesia,
para que uma freguesia não esteja representada apenas por um único local
de origem e destino das suas viagens (centróides).

Ao agregar as viagens todas em centróides, tornamos o exercício menos
realista pois elimina um conjunto importante de viagens de curta
distância, o que é uma característica de viagens em modos ativos
(\textbf{ref2?}). Para tal, recorreu-se ao método OD Jittering, que
utiliza uma desagregação de um único ponto (por exemplo o centróide de
uma área) em vários pontos aleatórios da rede viária existente, com base
no OpenStreetMap, e divide o volume de viagens dessa freguesia pelos
pares origem-destino gerados aleatoriamente.

Para este projeto, utilizou-se um nível de desagregação máxima de 100
viagens por par O-D. Isto significa que, por exemplo, para um par O-D de
2.000 viagens entre duas freguesias ligadas entre os seus centróides, o
método jittering dispersa aleatoriamente as 2.000 viagens em 20 pares de
100 viagens entre 20 origens e 20 destinos nas duas freguesias.

A figura 3 ilustra a diferença entre a representação de viagens
utilizando o método tradicional de ligação entre um único local entre
cada freguesia, e a representação com a aleatorização e desagregação de
viagens entre freguesias, para a área metropolitana de Lisboa. Figura 3
- Representação de pares OD na área metropolitana de Lisboa entre
freguesias, sem jittering (à esquerda), e com jittering (à direita).

O resultado da utilização deste método, em vez do tradicional uso de
centróides de grandes áreas, é que torna mais realistas a representação
de viagens realizadas, não correspondendo, no entanto, aos pares O-D
reais das viagens realizadas (por motivos de proteção de dados definidos
pelo INE). Para uma melhor compreensão do método utilizado, consultar
Jittering: A Computationally Efficient Method for Generating Realistic
Route Networks from Origin-Destination Data (Lovelace, Félix, and
Carlino 2022)

@@

A identificação dos percursos entre as origens e destinos (para as
linhas de desejo) são um aspeto fundamental da modelação da rede
ciclável nesta metodologia, uma vez que o IMob não identifica os
percursos realizados por cada viagem, sendo necessário estimá-los. A
respetiva modelação depende do cenário escolhido e tem em conta toda a
restante informação disponível (como os declives, ou localização de
interfaces de transporte público), para além da informação viária base.

A partir das linhas de desejo do cenário base, traçam-se os percursos
modelados para o modo ciclável com base na rede viária (equivalente à
afetação das viagens à rede, 4.º e último passo do modelo clássico dos 4
passos em sistemas de transportes). A modelação dos percursos cicláveis
depende das variáveis consideradas, assim como das restrições
pré-definidas. Estas restrições podem privilegiar aspetos como
velocidade e volumes baixos, rotas mais diretas, percursos menos
declivosos, entre outros, adequados à utilização da bicicleta, sendo que
o algoritmo segue uma avaliação que resulta da ponderação das variáveis
consideradas. Mesmo o percurso menos declivoso, mas com maior volume de
tráfego, poderá ser o indicado se a função assim o definir.

O algoritmo de escolha de percurso escolhido foi o do r5r que é um
algoritmo que permite muita flexibilidade nas configurações de tipo de
percursos estimados, nomeadamente por permitir estimar com probabilidade
e incerteza os percursos que usam transportes públicos (considerando os
seus horários). O r5r permite a identificação de percursos mais diretos
ou mais seguros, recorrendo para isso ao nível de stress de tráfego
(Level of Traffic Stress, ou LTS), que varia de 1 a 4, sendo o 1 mais
tranquilo - correspondendo, por exemplo, a pistas cicláveis fora de
estrada, e o 4 menos tranquilo - correspondendo, por exemplo, a
percursos que partilham o tráfego motorizado. Os percursos foram
estimados para o cenário base, para ambos os tipos de rede, direto e
seguro, usando para isso o LTS 4 e LTS 3 respetivamente.

O nível de tranquilidade\footnote{see
  \url{https://www.cyclestreets.net/help/journey/howitworks/\#quietness}}
é outro indicador estimado pelo CycleStreets para cada troço, com base
no número de vias, velocidade máxima permitida, arborização, hierarquia
e tráfego (quando a informação correspondente existe nas etiquetas do
OpenStreetMap). É apresentado numa escala de 0 a 100, em que o zero
corresponde ao nível menos seguro para se circular de bicicleta, e o 100
ao nível mais seguro e tranquilo - normalmente correspondendo a troços
que incluem já infraestrutura ciclável segregada do tráfego motorizado.

Foi utilizado um modelo digital do terreno de 25m de resolução espacial
e erro médio de altimetria vertical de 2.12m8 em Portugal, da Agência
Espacial Europeia, na sua missão COPERNICUS, para incluir impedâncias 9
de declives no modelo. O modelo r5r utiliza a rede viária do
OpenStreetMap (extraída a 22 de outubro 2022), e os dados de GTFS
criados e validados para o efeito. Foi estimado o número de viagens em
bicicleta potencial para as duas metas ENMAC (5\% e 10\%) a partir dos
valores de viagens em bicicleta e em automóvel (como condutor e como
passageiro) do cenário base de 2017.

Os percursos ou rotas foram depois sobrepostos e agregados por
segmentos, juntando a informação de várias rotas que se sobrepõem em
certos segmentos (overline). Por exemplo, foram somados os volumes de
viagens em bicicleta estimados e a transferência do automóvel, e foi
feita a média da velocidade automóvel e do nível de tranquilidade,
ponderadas pela distância dos segmentos.

@@

The intermodality scenario considers trips that can combine PT and
cycling for the first-and-last legs. Conservatively, we considered the
sum of first-and-last legs up to 5 km. Furthermore, we restricted PT use
to unimodal trips without transfers (although they can be included in
future modeling). Félix, Lovelace, and Moura (2022)

Finally, we only included PT modes that can practically accommodate
bicycles, such as trains, ferries, trams, and intermunicipal bus lines
with bike racks (~\ref{fig:map1}).

Foram recolhidos os dados de transportes públicos da AML, em formato
GTFS, tendo sido necessário o trabalho de limpeza, união, e validação
prévio. Esta informação é indispensável para saber que terminais de
transporte público (TP) se ligam a que origens ou destinos, e em que
horários, para não se considerar de forma generalista que uma interface
se liga a todo o sistema e que pode servir todas as viagens (origens e
destinos), o que não seria realista.

\begin{figure}
\includegraphics[width=0.6\linewidth,]{img/map1} \caption{Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area}\label{fig:map1}
\end{figure}

To obtain reliable results, we used the OpenStreetMap road network and
GTFS data. The r5r R package estimated the trip duration and distance
for both the original modes and the bicycle + PT combination, while the
od jittering R package estimated the OD locations based on a
centroid-based OD matrix.

Para a modelação dos percursos definiu-se como janela temporal viagens
até 120 minutos (2h). Para além disso considerou-se como duração máxima
25 minutos em bicicleta.

Map result (~\ref{fig:map2}).

\begin{figure}
\includegraphics[width=0.8\linewidth,]{img/map2} \caption{Bike routes with highest potential to serve as first and last mile when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
\end{figure}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
Félix, Rosa, Robin Lovelace, and Filipe Moura. 2022. {``{biclaR -
Ferramenta de apoio ao planeamento da rede ciclável na área
metropolitana de Lisboa}.''} {CERIS - Instituto Superior Técnico and
Transportes Metropolitanos de Lisboa}.
\url{https://biclar.tmlmobilidade.pt}.

\leavevmode\vadjust pre{\hypertarget{ref-bib2}{}}%
Slifka, M. K., and J. L. Whitton. 2000. {``Clinical Implications of
Dysregulated Cytokine Production.''} \emph{J. {M}ol. {M}ed.} 78: 74--80.
\url{https://doi.org/10.1007/s001090000086}.

>>>>

\usepackage{hyperref}
\hypersetup{colorlinks = TRUE,  urlcolor = blue, linkcolor = blue, citecolor = blue}

@@

\maketitle              % typeset the header of the contribution
%
\begin{abstract}
In metropolitan areas, car trips can be replaced by a combination of
public transit and cycling for the first-and-last mile. This paper
estimates the potential for cycling + public transit (PT) as a
substitute for car trips in the Lisbon metropolitan area and assesses
its socio-environmental impacts using open data and open source tools. A
decision support tool that facilitates the design and development of a
metropolitan cycling network was developed (\emph{biclaR}). A scenario
of intermodality introduced, and its socio-environmental impacts were
assessed using the \emph{HEAT for Cycling} and the \emph{HEAT as a
Service} tools. The impacts of shifting car trips to PT were also
estimated and monetized. The results indicate that 20\% of the current
trips can be made with the bicycle + PT combination. Shifting to cycling
for the first-and-last mile can reduce annual CO\textsubscript{2}eq
emissions from 6,000 tons/day, and the 10-year socio-environmental
benefits account from €230 million. For the PT leg, the transfer from
car avoids of at least 8,500 tons of CO\textsubscript{2}eq emissions per
year. The information on socio-economic benefits can support
policymakers in prioritizing interventions to reduce the reliance on
individual motorized transportation and effectively communicate their
decisions.

@@

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can significantly replace private car trips.
This approach requires interventions and programs to make bicycling more
appealing, and the resulting public investments can have significant
social and environmental benefits.

According to the latest mobility survey conducted in 2018 {[}1{]}, the
LMA registered a total of 5.3 million daily trips, with only 0.5\% by
bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%. The
number of intra-municipal trips - with origin and destination in the
same municipality - amounts to 3.5 million trips, exceeding the number
of inter-municipal trips (1.8 million trips) - involving travel between
different municipalities. Cars and public transport are the most used
modes for intercity trips, with cars being the predominant choice for
all journeys.

To achieve the cycling targets set by the Portuguese national cycling
strategy for 2025 and 2030 (4\% and 10\%, respectively) {[}2{]}, the
Lisbon's Metropolitan Department of Transport introduced
\emph{biclaR}\footnote{see
  \href{https://biclar.tmlmobilidade.pt/}{biclar.tmlmobilidade.pt}}, a
decision support tool that facilitates the design and development of a
metropolitan cycling network {[}3{]}.

\textbf{(RL)} The methods of Propensity to Cycle Tool\\
adding up an intermodality scenario to estimate cycling potential to
public transit interfaces.

This paper estimates the potential for combining cycling and PT to
substitute car trips in the LMA. After presenting the methods used, it
assesses its socio-environmental impacts using open data and open-source
tools.

@@

@@

The mobility survey data {[}1{]} is the basis for this project and
defines the baseline scenario. Despite being conducted in the
pre-pandemic period (2017), this dataset represents the most
comprehensive and up-to-date information on urban mobility in Portuguese
metropolitan areas (Lisbon and Porto).

We used a method for disaggregating the origins and destinations of
trips between the centroids of two districts (same as ``parish'') to
ensure that a district is not solely characterized by a single point of
origin and destination for its trips. Aggregating all trips into
centroids renders the exercise less realistic, as it excludes a
significant portion of short-distance trips, a prevalent characteristic
of active mode travel {[}4{]}. The OD Jittering method breaks down a
single point (i.e., the centroid of an area) into multiple random points
on the existing and neighboring road network, using OpenStreetMap as a
reference. This method then distributes the volume of trips within the
district among the randomly generated origin-destination pairs.

Using the
\href{https://github.com/dabreegster/odjitter}{\texttt{odjitter} R
package}, we employed a maximum disaggregation level of 100 trips per
O-D pair for this project. Figure \ref{fig:jitter} illustrates the
contrast between trip representation through the traditional method,
which connects a single desire line between each district, and the
presentation achieved through the randomization and disaggregation of
trips between districts, specifically for the Lisbon metropolitan area.

\begin{figure}

{\centering \includegraphics[width=1\linewidth,]{img/jitter}

}

\caption{Representation of OD pairs in the Lisbon metropolitan area between districts, without jittering (left) and with jittering (right).}\label{fig:jitter}
\end{figure}

Although this method provides a more realistic representation of the
trips undertaken compared to the traditional approach, it does not fully
align with the actual O-D pairs of trips, which remain unknown due to
data privacy regulations.

@@

The mobility survey collects the origin and destination of trips but
does not include the respective routes. Modeling the realistic cycling +
PT routes between OD pairs depends on assumptions regarding the
characteristics of the cycling and road networks and the location of
public transport interfaces. Other constraints regarding the behavior of
potential cyclists determine the routing results. For example, such
restrictions can favor low speed, low traffic streets, more direct
routes, and less steep paths, among others, suitable for cycling.

The selected route choice algorithm was the
\href{https://ipeagit.github.io/r5r/}{\texttt{r5r} R package} {[}5{]},
which allows for great flexibility in configuring estimated route types,
and which proven to provide most accurate route networks for the city of
Lisbon {[}6{]}. \texttt{r5r} can calculate multi-modal routes using PT
combined with other modes. It enables the identification of the most
direct or safest cycling routes, using the Level of Traffic
Stress\footnote{see
  \href{https://docs.conveyal.com/learn-more/traffic-stress}{docs.conveyal.com/learn-more/traffic-stress}}
(LTS) scale, ranging from 1 to 4, where 1 corresponds to the quietest
(e.g., off-road cycle paths) and 4 corresponds to the least quiet (e.g.,
routes shared with motorized traffic). The routes were estimated for the
base scenario for both types of networks: \emph{direct} and \emph{safe},
using LTS 4 and LTS 3, respectively.

The \texttt{r5r} model used the OpenStreetMap road network and the GTFS
metropolitan data agregated and validated. This information is crucial
for an accurate PT trip and route estimation. A digital elevation model
from the European Space Agency's COPERNICUS mission, with a 25m spatial
resolution, was used to include street gradient information, as a weight
in cycling routing. The cycling potential trips for the two ENMAC
targets (4\% and 10\%) was estimated from the values for cycling and car
trips (both as a driver and as a passenger) from the 2017 base scenario.

The routes were then overlaid and aggregated by segments, using
\href{https://docs.ropensci.org/stplanr/reference/overline.html}{\texttt{stplanr\ overline()}
R function}.

@@

The intermodality scenario considers trips combining PT and cycling for
the first and last legs. In a conservative approach, we have restricted
our analysis to the first and last legs with a combined length of up to
5 km or up to 25 minutes on bike. Furthermore, we have imposed
restrictions on PT usage to include only trips with no PT transfers, and
up to 2 hours (120 min). Additionally, we have only included PT modes
that can easily accommodate bicycles, such as trains, ferries, trams,
and inter-municipal bus lines equipped with bike racks (Figure
\ref{fig:map1}).

\begin{figure}

{\centering \includegraphics[width=0.6\linewidth,]{img/map1}

}

\caption{Interfaces and lines considered, by transport mode, in the Lisbon metropolitan area}\label{fig:map1}
\end{figure}

Figure \ref{fig:map2} illustrates the resulting bicycle routes to access
the main PT interfaces in the LMA.

\begin{figure}

{\centering \includegraphics[width=0.8\linewidth,]{img/map2}

}

\caption{Bike routes with highest potential to serve as first and last leg when replacing cycling and PT from car trips (screenshot of the interactive online tool).}\label{fig:map2}
\end{figure}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-IMOB}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{INE: Mobilidade e funcionalidade do território nas
{Áreas Metropolitanas do Porto e de Lisboa}: 2017,
\url{https://www.ine.pt/xportal/xmain?xpid=INE\&xpgid=ine_publicacoes\&PUBLICACOESpub_boui=349495406\&PUBLICACOESmodo=2\&xlang=pt},
(2018).}

\leavevmode\vadjust pre{\hypertarget{ref-ENMAC}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{Presidência do Conselho de Ministros: Resolução do
conselho de ministros n.º 131/2019,
\url{https://files.dre.pt/1s/2019/08/14700/0004600081.pdf}, (2019).}

\leavevmode\vadjust pre{\hypertarget{ref-felix2023}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{Félix, R., Lovelace, R., Moura, F.: {biclaR - Ferramenta
de apoio ao planeamento da rede ciclável na área metropolitana de
Lisboa}, \url{https://biclar.tmlmobilidade.pt}, (2022).}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022Jittering}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Jittering: A
computationally efficient method for generating realistic route networks
from origin-destination data. Findings. (2022).
https://doi.org/\href{https://doi.org/10.32866/001c.33873}{10.32866/001c.33873}.}

\leavevmode\vadjust pre{\hypertarget{ref-r5r}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga,
C.K.V., Conway, M.W.: r5r: Rapid realistic routing on multimodal
transport networks with R5 in r. Findings. (2021).
https://doi.org/\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}.}

\leavevmode\vadjust pre{\hypertarget{ref-Lovelace2022exploring}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Lovelace, R., Félix, R., Carlino, D.: Exploring
jittering and routing options for converting origin-destination data
into route networks: Towards accurate estimates of movement at the
street level. The International Archives of the Photogrammetry, Remote
Sensing and Spatial Information Sciences. XLVIII-4/W1-2022, 279--286
(2022).
https://doi.org/\href{https://doi.org/10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}{10.5194/isprs-archives-XLVIII-4-W1-2022-279-2022}.}
